SIMD-vectorized Chemical Source Term Evaluation

### 2017-02-23 paper.tex
first pass at dr. sungs edits

\begin{figure}[htb]
  \centering
  \begin{minipage}[t]{0.45\linewidth}
      \includegraphics[width=\textwidth]{SIMD.pdf}
      \caption{Schematic of SIMD processing.  A single processing element (e.g. a CPU core) contains a vector unit with several lanes (L0, L1, etc.).  The vector unit executes a single instruction concurrently on multiple data.}
      \label{F:SIMD}
  \end{minipage}
  \hfill
  \begin{minipage}[t]{0.45\linewidth}
      \includegraphics[width=\textwidth]{mem_access.pdf}
      \caption{Representation of an 2-D array with dimension $N_1 \times N_2$ in CPU memory, with linear index marked.}
      \label{F:mem}
  \end{minipage}
\end{figure}

@@

\subsection{Data ordering and vectorization patterns}
\label{S:data}
When storing arrays for a chemical kinetic model, e.g. the concentration of species $k$ for the $j$-th of thermo-chemical state $[C_k]_j$ there are two data-ordering choices.
In the ``F'' (Fortran column-major) format, the concentrations of a single species $k$ over all $N$ thermo-chemical states are adjacent.
In Fig.~\ref{F:mem} this would correspond to storing $[C_0]_0$ in index \num{0}, $[C_0]_1$ in index \num{1} and so on, with $N_1 = N$, and $N_2 = N_s$, the number of species in the model.
In the ``C'' (C row-major) format, the storage order is transposed with $[C_0]_0$ in index \num{0}, $[C_1]_0$ in index \num{1} etc., and $N_1 = N_s$, $N_2 = N$.
This ordering---along with the device (CPU, GPU, etc.) in question---have a large effect on the performance of SIMD\slash SIMT-vectorized algorithms.

In a \textit{shallow}-SIMD vectorization, each SIMD-lane evaluates chemical source terms for a different thermo-chemical state.
If the data is stored in ``F''-order, the SIMD-lanes accessing the concentration of species $k$ for states $j, j+1\ldots$ will load sequential locations in memory, however the $k+1$th species concentration will be $N$ memory locations away increasing the likelihood of cache-misses on the CPU.
In a \textit{deep}-SIMD vectorization, a vector-unit cooperates to evaluate chemical source terms for a single thermo-chemical state $j$, thus SIMD-lanes loading species concentrations $k, k+1\ldots$ will access sequential memory locations if the data is stored in ``C''-order.
Further, in the ``C''-ordering the furthest difference between any two concentrations in a single thermo-chemical state is at most $N_s$, with $N_s \ll N$ in most cases; this greatly improved data-locality increases the chances of a cache-hit on the CPU.
However, a deep vectorization may result in SIMD-\textit{waste}---similar to thread-divergence in GPU-processing---caused by different SIMD-lanes executing different instructions (e.g. from differing for-loop bounds).

Finally, in a SIMT-vectorization, threads execute the same instruction over multiple data in lock-step without use of specialized vectorization instructions.
This is the paradigm used in GPU-processing, a more through review of which can be found in our previous work~\cite{CurtisGPU:2017}; one key-exception in this case is that the algorithm executed to evaluate the chemical source terms is identical for all thermo-chemical states, thus there is no thread-divergence for a shallow SIMT-vectorization. 
In Sec.~\ref{S:results} the effect of the data-ordering on the shallow-SIMD\slash SIMT vectorization performance will be studied.

\subsection{Previous works and goals of this study}
Several recent works, e.g. \cite{CurtisGPU:2017} have investigated SIMT-based chemical kinetic integration on GPUs, however SIMD-based chemical kinetics have been less studied in comparison.
\textcite{stone2016} implemented a linearly-implicit fourth-order stiff Rosenbrock solver in the OpenCL for various platforms including the CPU, GPU and the MIC.
The shallow-SIMD and SIMT-vectorization models were implemented in OpenCL and compared to an OpenMP baseline code that was deep SIMD-vectorized by simple compiler hints (a.k.a pragmas).
The shallow-SIMD vectorization improved the integrator performance over the OpenMP baseline by \SIrange{2.5}{2.8}{$\times$} on the CPU and \SIrange{4.7}{4.9}{$\times$} on the MIC, while the GPU performance was \SIrange{1.4}{1.6}{$\times$} slower than the OpenMP baseline due to thread-divergence concerns.
\textcite{kroshko2013efficient} implemented a shallow-SIMD vectorized third order stiff Rosenbrock integrator for atmospheric chemistry on a Cell Broadband Engine---a specially designed vector processor---finding a speedup of \SI{1.89}{$\times$} over a serial version of the same code, a \SI{94}{\percent} parallel efficiency.

@@

\begin{equation}
\frac{\text{d}[C_k]}{\text{d}t} = \sum_{i}^{N_R} \nu_{k,i} R_i c_i
\end{equation}
where $\frac{\text{d}[C_k]}{\text{d}t}$, $N_R$, $\nu_{k,i}$, ${R_i}$, $c_i$ are the time rate of change of the concentration of species $k$, the number of reactions in the model, the net stoichiometric coefficient of species $k$ in reaction $i$, the net rate of progress of reaction $i$, and the pressure modification of reaction $i$ respectively.
For further detail on each term, the reader is referred to our previous work~\cite{Niemeyer:2016aa}; the code is capable of evaluating all modern reaction rate types (e.g. pressure-log, chebyshev, etc.).

In addition, the temperature rate of change using the constant-pressure assumption\footnote{Note: in this context, the ``constant-pressure assumption'' refers to evaluation within a reaction sub-step in the operator splitting scheme, rather than a general constant-pressure reactive-flow simulation.} is evaluated as:
\begin{equation}
\frac{\text{d}[T]}{\text{d}t} = -\frac{\sum_{k}^{N_S} H^{\degree}_k \frac{\text{d}[C_k]}{\text{d}t}}{\sum_{k}^{N_S} [C_k] c_{p_k}^{\degree}}
\end{equation}
where $H^{\degree}_k$ and $c_{p_k}^{\degree}$ are the molar specific enthalpy and specific heat respectively.
Although the code is equally capable of a evaluating the temperature rate using a constant-volume assumption, we omit this here for brevity.

@@

\section{Results and Discussion}
\subsection{Validation}
The reaction rates of progress (ROP), species and temperature rates in this study are validated by comparison with Cantera~\cite{Cantera}, however special care must be taken due floating point arithmetic issues

In computing the net ROP of reaction $i$ from the forward and reverse ROP: $R_{i} = R_{i}^{\prime} - R_{i}^{\prime\prime}$, precision can easily be lost as the net ROP may be---particularly near chemical equilibrium---many orders of magnitude smaller than the forward or reverse rates.
A base-ten bound on loss of precision~\cite{goldberg1991every} can be found as:
\begin{equation}
\text{L}_{i,\text{min}} = 2^{-p_i} \le \left\lvert 1 - \frac{R_{i}^{\prime}}{R_{i}^{\prime\prime}} \right\rvert \le 2^{-q_i} = \text{L}_{i,\text{max}}}
\label{e:bounds}
\end{equation}
with $q_i \le p_i$.
To obtain an estimate of the precision bounds, the forward and reverse ROP from Cantera are substituted into Eq.~\eqref{e:bounds}.
Additionally a higher-precision number format (in this case, the Python Decimal class) is used in this computation to maintain accuracy.
Finally, when determining the error in the net ROP between this code and Cantera, the bounds on the percent error attributable to precision loss, $\text{P}_{i,\text{err}}$ can be estimated as:
\begin{equation}
\text{P}_{i,\text{min}} = 100 * \frac{\text{L}_{i,\text{min}}}{\left\lvert\ R_{i,\text{CT}} - R_{i}\right\rvert} \le
\text{P}_{i,\text{err}} \le 100 * \frac{\text{L}_{i,\text{max}}}{\left\lvert R_{i,\text{CT}} - R_{i} \right\rvert} = \text{P}_{i,\text{max}}
\label{e:rel_bound}
\end{equation}
where the \text{CT} subscript indicates values from Cantera.

For a more direct comparison, a relative error norm of a quantity $X_{i,j}$ over all states $j$, and indices $i$ was computed using the $L^{\infty}$ norm:
\begin{equation}
 E_{X_{i}} = \left\lVert \frac{\left\lvert X_{i,j,\text{CT}} - X_{i,j}\right\rvert}{\num{e-10} + \num{e-6} * \left\lvert X_{i,j,\text{CT}} \right\rvert} \right\rVert_{i,j,\infty}
\label{e:rel_err}
\end{equation}
and the lower precision bound for the net ROP error norm is set to $\text{P}_{\text{min}}^{*} = \text{P}_{i,\text{min}}$ for the $i, j$ value that results in the maximum error $E_{R_i}$.

\begin{table}[htb]
\sisetup{retain-zero-exponent=true}
\centering
\begin{tabular}{@{}S[table-format=1.2e1] S[table-format=1.2e1] S[table-format=1.2e1] S[table-format=1.2e1] S[table-format=1.2e1] S[table-format=1.2e1] S[table-format=1.2e1] @{}}
\toprule
\multicolumn{1}{l}{Model} & \multicolumn{1}{c}{$E_{R_{i}^{\prime}}$} & \multicolumn{1}{c}{$E_{R_{i}^{\prime\prime}}$} &\multicolumn{1}{c}{$E_{R_{i}}$} & \multicolumn{1}{c}{$\text{P}_{\text{min}}^{*}\si{\percent}$} & \multicolumn{1}{c}{$E_{\frac{\text{d}[C_k]}{\text{d}t}}$} & \multicolumn{1}{c}{$E_{\frac{\text{d}[T]}{\text{d}t}}$} \\
\midrule
\multicolumn{1}{l}{\ce{H2}\slash~\ce{CO}\cite{Burke:2011fh}} & 1.20e-8 & 6.33e-8 & 1.32e1 & 4.66e6 & 2.37e1 & 2.12e5 \\
\multicolumn{1}{l}{GRI-Mech.~3.0\cite{smith_gri-mech_30}}  & 3.07e-8 & 5.90e-8 & 1.21e0 & 2.67e04 & 2.84e0 & 2.64e4 \\
\multicolumn{1}{l}{USC-Mech II\cite{Wang:2007}}  & 1.07e-7 & 1.37e-7 & 2.39e0 & 4.12e02 & 7.92e00 & 3.22e03 \\
\multicolumn{1}{l}{\ce{IC5H11OH}\cite{Sarathy:2013jr}} & -1 &  -1 & -1 & -1 & -1 & -1 \\
\bottomrule
\end{tabular}
\caption{Summary of rate of progress, species and temperature rate correctness.
Error statistics are based on the infinity-norm of the relative error detailed in Eq.~\eqref{e:rel_err} for each quantity.
\todo[inline]{i-pentanol validation still running}
}
\label{T:error}
\end{table}

In Table~\ref{T:error}, we see the results of this code as compared to Cantera on a library of partially stirred reaction conditions (PaSR) described in our previous works~\cite{CurtisGPU:2017,Niemeyer:2016aa}.
Very close agreement is observed for the forward and reverse ROP for all models, however the net ROP error norm is \numrange{7}{8} orders of magnitude larger.
We see here that the minimum estimated loss of precision is \SIrange{4}{40}{$\times$} the error in net ROP, explaining this jump.
The species rate error norm is similar in magnitude to that of the net ROP, however the temperature rate error again increases as it depends directly on the species rates.
We note that the above discussion does not imply that computation of the net ROP will cause large error in chemical kinetic integration---either in this code or Cantera---as this loss of precision only occurs when the forward and reverse ROP are nearly equal, implying the reaction is in near-equilibrium.

\subsection{Results}
\label{S:results}
The performance studies in this work were run on four ten-core \SI{2.2}{\giga\hertz} Intel Xeon E5-4640 v2 CPUs with \SI{20}{\mega\byte} of L3 cache memory, installed on an Ace Powerworks PW8027R-TRF+ with a Supermicro X9QR7-TF+/X9QRi-F+ baseboard.
Runtimes in each case were averaged over ten runs, each using the same set of PaSR conditions utilized in validation.
All code was compiled with \texttt{gcc 4.8.5} and ``\texttt{-O3}'' optimization, running on \texttt{v16.1.1} of the Intel OpenCL runtime supporting OpenCL \texttt{v1.2}.

@@

The effect of changing the vector-width supplied to a ``C''-ordered shallow-SIMD vectorization on a single CPU core is investigated in Fig.~\ref{F:Veccomp}.
Significant speedups are seen with larger vector-widths for the smaller models, however as the model size increases data must be loaded from increasingly distant memory locations and the performance of all three vector-widths becomes similar.
Finally in Fig.~\ref{F:SIMDscale} the strong parallel scaling of the chemical source term evaluation is examined for all four models with a ``C''-ordered shallow SIMD-vectorization and vector-width of 16:
Interestingly the efficiency for the hydrogen model decreases fastest, while the larger models have similar efficiencies; the isopentanol model has the highest efficiency in all cases.
This suggests that evaluation of the chemical-source terms for a single thermo-chemical state for the smaller models may not saturate the processing power of the CPU, some efficiency may be lost due to context switching within OpenCL itself.
Additionally, OpenCL allows assigning of work to devices by shared memory resources e.g., a L3 cache, which was not pursued in this study for simplicity; this should be investigated in future work to determine the effect on scaling efficiency.

\section{Conclusions}
In this work, a shallow SIMD-vectorized chemical source term evaluation OpenCL code generator was developed and validated against Cantera~\cite{Cantera} for a wide range of chemical kinetic models~\cite{Burke:2011fh,smith_gri-mech_30,Wang:2007,Sarathy:2013jr}.
Significant speedups of up to \SIrange{1.88}{2.42}{$\times$} over a baseline SIMT-vectorized code were observed.
Two data-ordering schemes were investigated, showing a clear performance benefit for use of the ``C'' (row-major) ordering for the shallow vectorized code.
Further, this study suggests that a deep SIMD-vectorized code---currently under development---may see even greater accelerations due to expected increases in cache-hit rates.
Larger vector-widths were found to provide accelerations for the smaller models studied, but performance was similar over all vector-widths tested for the largest models tested.
Finally, the strong parallel scaling efficiency was examined for the shallow vectorized code; the larger models exhibited higher scaling efficiency, possibly due to failure of the \ce{H2}\slash\ce{CO} model to saturate the CPU throughput.
Future extensions of this work will include development of a SIMD\slash SIMT-accelerated sparse-analytical Jacobian code targeted at CPUs, GPUs and MICs.

\section{Acknowledgements}
This research was funded by \ldots

>>>>

\begin{figure}[htb]
  \centering
  \begin{minipage}[t]{0.45\linewidth}
      \includegraphics[width=\textwidth]{SIMD.pdf}
      \caption{Schematic of SIMD processing.  A single processing element (e.g. a CPU core) contains a vector unit with $N$ lanes ($\text{L}_0,\ldots \text{L}_N$).  The vector unit executes a single instruction concurrently on multiple data.}
      \label{F:SIMD}
  \end{minipage}
  \hfill
  \begin{minipage}[t]{0.45\linewidth}
      \includegraphics[width=\textwidth]{mem_access.pdf}
      \caption{Representation of an 2-D array with dimension $N \times K$ in CPU memory, with linear index marked.}
      \label{F:mem}
  \end{minipage}
\end{figure}

@@

\subsection{Data ordering and vectorization patterns}
\label{S:data}
When storing arrays for a chemical kinetic model, e.g. the concentration of species $k$ for the $j$-th of thermo-chemical state $[C_k]_j$---with $1 \le k \le N_{\text{sp}}$ the number of species in the model, and $1 \le j \le N_{\text{state}}$ the number of thermo-chemical states considered---there are two data-ordering choices.
In the ``F'' (Fortran column-major) format, the concentrations of a single species $k$ over all thermo-chemical states are adjacent.
In Fig.~\ref{F:mem} this would correspond to storing $[C_1]_1$ in index \num{0}, $[C_1]_2$ in index \num{1} and so on, with $N = N_{\text{state}}$, and $K = N_{\text{sp}}$.
In the ``C'' (C row-major) format, the storage order is transposed with $[C_1]_1$ in index \num{0}, $[C_2]_1$ in index \num{1} etc., and $N = N_{\text{sp}}$, $K = N_{\text{state}}$.
This ordering---along with the device (CPU, GPU, etc.) in question---have a large effect on the performance of SIMD\slash SIMT-vectorized algorithms.

In a \textit{shallow}-SIMD vectorization, each SIMD-lane evaluates chemical source terms for a different thermo-chemical state.
If the data is stored in ``F''-order, the SIMD-lanes accessing the concentration of species $k$ for states $1, 2\ldots$ will load sequential locations in memory, however the $(k+1)$th species concentration will be $N_{\text{state}}$ memory locations away increasing the likelihood of cache-misses on the CPU.
In a \textit{deep}-SIMD vectorization, a vector-unit cooperates to evaluate chemical source terms for a single thermo-chemical state $j$, thus SIMD-lanes loading species concentrations $k, k+1\ldots$ will access sequential memory locations if the data is stored in ``C''-order.
Further, in the ``C''-ordering the furthest difference between any two concentrations in a single thermo-chemical state is at most $N_{\text{sp}}$, with $N_{\text{sp}} \ll N_{\text{state}}$ in most cases; this greatly improved data-locality increases the chances of a cache-hit on the CPU.
However, a deep vectorization may result in SIMD-\textit{waste}---similar to thread-divergence in GPU-processing---caused by different SIMD-lanes executing different instructions (e.g. from differing for-loop bounds).

Finally, in a SIMT-vectorization, threads execute the same instruction over multiple data in lock-step without use of specialized vectorization instructions.
This is the paradigm used in GPU-processing, a more thorough review of which can be found in our previous work~\cite{CurtisGPU:2017}; one key-exception in this case is that the algorithm executed to evaluate the chemical source terms is identical for all thermo-chemical states, thus there is no thread-divergence for a shallow SIMT-vectorization. 
In Sec.~\ref{S:results} the effect of the data-ordering on the shallow-SIMD\slash SIMT vectorization performance will be studied.

\subsection{Previous works and goals of this study}
Several recent works, e.g. \cite{CurtisGPU:2017}, have investigated SIMT-based chemical kinetic integration on GPUs, however SIMD-based chemical kinetics have been less studied in comparison.
\textcite{stone2016} implemented a linearly-implicit fourth-order stiff Rosenbrock solver in the OpenCL for various platforms including CPUs, GPUs, and MICs.
The shallow-SIMD and SIMT-vectorization models were implemented in OpenCL and compared to an OpenMP baseline code that was deep SIMD-vectorized by simple compiler hints (a.k.a pragmas).
The shallow-SIMD vectorization improved the integrator performance over the OpenMP baseline by \SIrange{2.5}{2.8}{$\times$} on the CPU and \SIrange{4.7}{4.9}{$\times$} on the MIC, while the GPU performance was \SIrange{1.4}{1.6}{$\times$} slower than the OpenMP baseline due to thread-divergence concerns.
\textcite{kroshko2013efficient} implemented a shallow-SIMD vectorized third order stiff Rosenbrock integrator for atmospheric chemistry on a Cell Broadband Engine---a specially designed vector processor---finding a speedup of \SI{1.89}{$\times$} over a serial version of the same code.

@@

\begin{equation}
\dot{\omega}_k = \sum_{i=1}^{N_{\text{reac}}} \nu_{ki} R_i c_i
\end{equation}
where $\dot{\omega}_k$, $N_{\text{reac}}$, $\nu_{k,i}$, ${R_i}$, $c_i$ are the time rate of change of the concentration of species $k$, the number of reactions in the model, the net stoichiometric coefficient of species $k$ in reaction $i$, the net rate of progress of reaction $i$, and the pressure modification of reaction $i$, respectively.
For further detail on each term, the reader is referred to our previous work~\cite{Niemeyer:2016aa}; the code is capable of evaluating all modern reaction rate types (e.g. pressure-log, Chebyshev, etc.).

In addition, the temperature rate of change using the constant-pressure assumption\footnote{Note: in this context, the ``constant-pressure assumption'' refers to evaluation within a reaction sub-step in the operator splitting scheme, rather than a general constant-pressure reactive-flow simulation.} is evaluated as:
\begin{equation}
\frac{\partial T}{\partial t} = -\frac{\sum_{k=1}^{N_{\text{sp}}} H^{\degree}_k \dot{\omega}_k}{\sum_{k=1}^{N_{\text{sp}}} [C_k] c_{p_k}^{\degree}}
\end{equation}
where $H^{\degree}_k$ and $c_{p_k}^{\degree}$ are the molar specific enthalpy and specific heat, respectively.
Although the code is equally capable of a evaluating the temperature rate using a constant-volume assumption, we omit this here for brevity.

@@

\section{Results and Discussion}
\subsection{Comparison to Cantera}
The reaction rates of progress (ROP), species and temperature rates in this study are validated by comparison with Cantera~\cite{Cantera}, however special care must be taken due floating point arithmetic issues

In computing the net ROP of reaction $i$ for state $j$ from the forward and reverse ROP: $R_{ij} = R_{ij}^{\prime} - R_{ij}^{\prime\prime}$, precision can easily be lost as the net ROP may be---particularly near chemical equilibrium---many orders of magnitude smaller than the forward or reverse rates.
A base-ten bound on loss of precision~\cite{goldberg1991every} can be found as:
\begin{equation}
\text{L}_{ij,\text{lower}} = 2^{-p_i} \le \left\lvert 1 - \frac{R_{ij}^{\prime}}{R_{ij}^{\prime\prime}} \right\rvert \le 2^{-q_i} = \text{L}_{ij,\text{upper}}
\label{e:bounds}
\end{equation}
with $q_i \le p_i$.
To obtain an estimate of the precision bounds, the forward and reverse ROP from Cantera~\cite{Cantera} are substituted into Eq.~\eqref{e:bounds}.
Additionally a higher-precision number format (in this case, a quad-precision floating point) is used in this computation to maintain accuracy.
Finally, when determining the error in the net ROP between this code and Cantera~\cite{Cantera}, the bounds on the percent error attributable to precision loss, $\text{P}_{i,\text{err}}$ can be estimated as:
\begin{equation}
\text{P}_{ij,\text{lower}} = 100 * \frac{\text{L}_{ij,\text{lower}}}{\left\lvert\ R_{ij,\text{CT}} - R_{ij}\right\rvert} \le
\text{P}_{ij,\text{err}} \le 100 * \frac{\text{L}_{ij,\text{upper}}}{\left\lvert R_{ij,\text{CT}} - R_{ij} \right\rvert} = \text{P}_{ij,\text{upper}}
\label{e:rel_bound}
\end{equation}
where the \text{CT} subscript indicates values from Cantera~\cite{Cantera}.
In the case that $R_{ij,\text{CT}} = R_{ij}$, the error of this net ROP is zero and the precision loss is ignored.

For a more direct comparison, a relative error norm of a quantity $X_{ij}$ over all states $j$, and reactions $i$ was computed using the $L^{\infty}$ norm:
\begin{equation}
 E_{X} = \left\lVert \frac{\left\lvert X_{ij,\text{CT}} - X_{ij}\right\rvert}{\num{e-10} + \num{e-6} * \left\lvert X_{i,j,\text{CT}} \right\rvert} \right\rVert_{i,j,\infty}
\label{e:rel_err}
\end{equation}
For the reaction $i$ and state $j$ that results in the maximum ROP error $E_{R}$, the lower precision bound is set to $\text{P}_{\text{lower}}^{*} = \text{P}_{i,\text{lower}}$.

\begin{table}[htb]
\sisetup{retain-zero-exponent=true}
\centering
\begin{tabular}{@{}S[table-format=1.2e1] S[table-format=1.2e1] S[table-format=1.2e1] S[table-format=1.2e1] S[table-format=1.2e1] S[table-format=1.2e1] S[table-format=1.2e1] @{}}
\toprule
\multicolumn{1}{l}{Model} & \multicolumn{1}{c}{$E_{R^{\prime}}$} & \multicolumn{1}{c}{$E_{R^{\prime\prime}}$} &\multicolumn{1}{c}{$E_{R}$} & \multicolumn{1}{c}{$\text{P}_{\text{lower}}^{*}\si{\percent}$} & \multicolumn{1}{c}{$E_{\dot{\omega}}$} & \multicolumn{1}{c}{$E_{\frac{\partial T}{\partial t}}$} \\
\midrule
\multicolumn{1}{l}{\ce{H2}\slash~\ce{CO}\cite{Burke:2011fh}} & 1.20e-8 & 6.33e-8 & 1.32e1 & 4.66e6 & 2.37e1 & 2.12e5 \\
\multicolumn{1}{l}{GRI-Mech.~3.0\cite{smith_gri-mech_30}}  & 3.07e-8 & 5.90e-8 & 1.21e0 & 2.67e04 & 2.84e0 & 2.64e4 \\
\multicolumn{1}{l}{USC-Mech II\cite{Wang:2007}}  & 1.07e-7 & 1.37e-7 & 2.39e0 & 4.12e02 & 7.92e00 & 3.22e03 \\
\multicolumn{1}{l}{\ce{iC5H11OH}\cite{Sarathy:2013jr}} & -1 &  -1 & -1 & -1 & -1 & -1 \\
\bottomrule
\end{tabular}
\caption{Summary of rate of progress, species and temperature rate correctness.
Error statistics are based on the infinity-norm of the relative error detailed in Eq.~\eqref{e:rel_err} for each quantity.
\todo[inline]{i-pentanol validation still running}
}
\label{T:error}
\end{table}

In Table~\ref{T:error}, we see the results of this code as compared to Cantera~\cite{Cantera} on a library of partially stirred reaction conditions (PaSR) described in our previous works~\cite{CurtisGPU:2017,Niemeyer:2016aa}.
Very close agreement is observed for the forward and reverse ROP for all models, however the net ROP error norm is \numrange{7}{8} orders of magnitude larger.
We see here that the estimated lower bound on loss of precision is \SIrange{4}{40}{$\times$} the error in net ROP, explaining the large increase in the net ROP error norm as compared to either the forward or backward ROP error norms.
The species rate error norm is similar in magnitude to that of the net ROP, however the temperature rate error again increases as it depends directly on the species rates.
We note that the above discussion does not imply that computation of the net ROP will cause large error in chemical kinetic integration---either in this code or Cantera~\cite{Cantera}---as this loss of precision only occurs when the forward and reverse ROP are nearly equal, implying the reaction is in near-equilibrium.

\subsection{Performance}
\label{S:results}
The performance studies in this work were run on four ten-core \SI{2.2}{\giga\hertz} Intel Xeon E5-4640 v2 CPUs with \SI{20}{\mega\byte} of L3 cache memory, installed on an Ace Powerworks PW8027R-TRF+ with a Supermicro X9QR7-TF+/X9QRi-F+ baseboard.
Runtimes in each case were averaged over ten runs, each using the same set of PaSR conditions utilized in validation.
All code was compiled with \texttt{gcc 4.8.5} and ``\texttt{-O3}'' optimization, running on \texttt{v16.1.1} of the Intel OpenCL runtime supporting OpenCL \texttt{v1.2}.

@@

The effect of changing the vector-width supplied to a ``C''-ordered shallow-SIMD vectorization on a single CPU core is investigated in Fig.~\ref{F:Veccomp}.
Significant speedups are seen with larger vector-widths for the smaller models, however as the model size increases data must be loaded from increasingly distant memory locations and the performance of all three vector-widths becomes similar.
Finally, the strong parallel scaling efficiency:
\begin{equation}
S = \frac{t_1}{N t_N}
\end{equation}
where $t_1$ is the mean wall-clock runtime of the chemical source terms evaluation on one core and $t_N$ the mean runtime on $N$ cores, is examined in Fig~.\ref{F:SIMDscale} for all four models with a ``C''-ordered shallow SIMD-vectorization and vector-width of 16.
Interestingly the efficiency for the hydrogen model decreases fastest, while the larger models have nearly identical for more than two cores; the isopentanol model has the highest efficiency in all cases.
This suggests that evaluation of the chemical-source terms for a single thermo-chemical state for the smaller models may not saturate the processing power of the CPU, some efficiency may be lost due to context switching within OpenCL itself.
Additionally, OpenCL allows assigning of work to devices by shared memory resources, e.g., a L3 cache, which was not pursued in this study for simplicity; this should be investigated in future work to determine the effect on scaling efficiency.

\section{Conclusions}
In this work, automatically generated OpenCL codes for shallow SIMD-vectorized chemical source term evaluation were developed and validated against Cantera~\cite{Cantera} for a wide range of chemical kinetic models~\cite{Burke:2011fh,smith_gri-mech_30,Wang:2007,Sarathy:2013jr}.
Significant speedups of up to \SIrange{1.88}{2.42}{$\times$} over a baseline SIMT-vectorized code were observed.
Two data-ordering schemes were investigated, showing a clear performance benefit for use of the ``C'' (row-major) ordering for the shallow vectorized code.
Further, this study suggests that a deep SIMD-vectorized code---currently under development---may see even greater accelerations due to expected increases in cache-hit rates.
Larger vector-widths were found to provide accelerations for the smaller models studied, but the speedup was similar over all vector-widths tested for the largest models investigated.
Finally, the strong parallel scaling efficiency was examined for the shallow vectorized code; the larger models exhibited higher scaling efficiency, possibly due to failure of the \ce{H2}\slash\ce{CO} model to saturate the CPU throughput.
Future extensions of this work will include development of a SIMD\slash SIMT-accelerated sparse-analytical Jacobian code targeted at CPUs, GPUs, and MICs.

\section{Acknowledgements}
This research was funded by the National Science Foundation under grant ACI-1534688.
